---
title: "**Analytic pipeline of DNA methylation data in R code**"
author: "**Irene D'Onofrio**"
date: "28th June 2023"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 2
    number_sections: false
    theme: lumen
    fig_retina: 4
  github_document:
    toc: true
    toc_depth: 2
    number_sections: false
editor_options: 
  chunk_output_type: inline
---

------------------------------------------------------------------------

## 1. Load raw data

Loading of raw data with minfi and creation of an object called RGset which stores the RGChannelSet object

```{r message=FALSE, warning=FALSE}
rm(list=ls())
setwd('C:/Users/Irene/Desktop/LM Bioinformatics/DRD/Module2/Final Report-20230616/')
library(minfi)

baseDir <- ("C:/Users/Irene/Desktop/LM Bioinformatics/DRD/Module2/Final Report-20230616/Input/")
targets <- read.metharray.sheet(baseDir)
RGset <- read.metharray.exp(targets = targets) 
save(RGset,file="RGset.RData")


```

## 2. Create R/G dataframes

Creation of the Red and Green dataframes to store the red and green fluorescences respectively

```{r message=FALSE, warning=FALSE}
Red <- data.frame(getRed(RGset))
dim(Red)
head(Red)

Green <- data.frame(getGreen(RGset))
dim(Green)
head(Green)

```

## 3. Check probe info by address

Retrieving the Red and Green fluorescences and the chemistry of the probes related to the following address: **64638362**. To check for the type of probes it is necessary to look into the Manifest file, which has been downloaded at the following from the from the Illumina website (<http://support.illumina.com/array/array_kits/infinium_humanmethylation450_beadchip_kit/downloads.html>) and then cleaned up (control and rs probes have been removed).

```{r message=FALSE, warning=FALSE}
red_probe <- Red[rownames(Red) == "64638362", ]
green_probe <- Green[rownames(Green)=="64638362",]

load("C:/Users/Irene/Desktop/LM Bioinformatics/DRD/Module2/DRD_2023_Lesson_2-20230516/Illumina450Manifest_clean.RData")
Illumina450Manifest_clean[Illumina450Manifest_clean$AddressA_ID=="64638362", 'Infinium_Design_Type']


df_address <- data.frame(Sample=colnames(green_probe), Red_fluor=unlist(red_probe, use.names = FALSE ), Green_fluor=unlist(green_probe, use.names = FALSE ), Type = "II")
df_address

```

It is possible to see that the address corresponds to a Type II probe thus no color channel is specified. Moreover, this kind of probes has a correspondence only for the address A and not for the address B.

## 4. Create the object MSet.raw

```{r message=FALSE, warning=FALSE}


MSet.raw <- preprocessRaw(RGset)
MSet.raw


```

## 5. Quality check

Quality checks:

1)  QCplot

2)  check of the intensity of negative controls using minfi

3)  calculation of the detection pValues; for each sample, how many probes have a detection p-value higher than the threshold **0.05**?

#### 5.1 QCplot

In a QCplot, the medians from the methylation signal and unmethylation signal distributions are plotted. Data of **good quality** are found at **high values of median for both the distributions** (that is right and upper part); on the contrary, low values of median indicate a lower quality of data.

The QC plot has two main limits:

-   It doesn't take into account the background signal

-   It doesn't take into account whether some failure happens during the sample preparation

The getQC() function creates a DataFrame with two columns: mMed and uMed which are the chipwide medians of the Meth and Unmeth channels. Through plotQC(), a diagnostic QC plot is produced from this dataframe.

```{r, fig.align = 'center',fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
qc <- getQC(MSet.raw)
plotQC(qc)
```

Here, it is clearly visible that **all the samples** cluster together with a good median value, indicating their **good quality**.

#### 5.2 Negative control intensity check

The negative controls probes are used to estimate the **background intensity** of the system. Under normal experimental condition, the intensity values for these probes can vary from 100 up to 1000 units in a sample-dependent way. An increase of a given sample may indicate poor quality of the DNA template prior to bisulphite conversion.

The check is performed through the function controlStripPlot(), that produces strip plots for the specified control probe type (only negative, in this analysis).

```{r, fig.align = 'center',fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
controlStripPlot(RGset, controls="NEGATIVE")
```

Here, in both channel the intensity values are below 10, thus they are in the acceptable range.

#### 5.3 Failed position check

The p-value can help to decide if you need to remove a specific sample, since it indicates the chance that the target sequence signal is distinguishable from the negative controls.

Usually, **bad sample** contains **more than 5%** of bad probes.

```{r, message=FALSE, warning=FALSE}
detP <- detectionP(RGset)
failed <- detP>0.05


num_failed = colSums(failed)


df_failed <- data.frame(Sample=colnames(green_probe), Failed_Position=num_failed, perc_failed_probes=(means_of_columns <- colMeans(failed))*100, row.names = NULL)

df_failed
```

None of the samples has a percentage of failed probes higher than 5%, thus we can retain them.

## 6. Beta and M values

Calculation of raw beta and M values and plotting of the densities of mean methylation values, dividing the samples in WT and MUT.

Beta and M values are two metrics to measure methylation levels. The M-value is more statistically valid for the differential analysis of methylation levels. However, the β-value is much more biologically interpretable.

```{r, message=FALSE, warning=FALSE}
wt <- targets[targets$Group=="WT", "Basename"]
mut <- targets[targets$Group=="MUT", "Basename"]
wt <- gsub(baseDir, "", wt)
mut <- gsub(baseDir, "", mut)

wtSet <- MSet.raw[,colnames(MSet.raw) %in% wt]
mutSet <- MSet.raw[,colnames(MSet.raw) %in% mut]

wtBeta <- getBeta(wtSet)
wtM <- getM(wtSet)
mutBeta <- getBeta(mutSet)
mutM <- getM(mutSet)

mean_wtBeta <- apply(wtBeta,MARGIN=1,mean,na.rm=T)
mean_mutBeta <- apply(mutBeta,MARGIN=1,mean,na.rm=T)
mean_wtM <- apply(wtM,MARGIN=1,mean,na.rm=T)
mean_mutM <- apply(mutM,MARGIN=1,mean,na.rm=T)


d_mean_wtBeta <- density(mean_wtBeta,na.rm=T)
d_mean_mutBeta <- density(mean_mutBeta,na.rm=T)
d_mean_wtM <- density(mean_wtM,na.rm=T)
d_mean_mutM <- density(mean_mutM,na.rm=T)
```

```{r, fig.align = 'center',fig.height=8, fig.width=16, message=FALSE, warning=FALSE}

par(mfrow=c(1,2))
plot(d_mean_wtBeta,main="Density of Beta Values",col="#0067E6", lwd=2.5)
lines(d_mean_mutBeta,main="Density of Beta Values",col="#E50068", lwd=2.5)

legend('topright', legend=c("WT", "MUT"), 
       fill = c("#0067E6","#E50068"), cex=1
)
plot(d_mean_wtM,main="Density of M Values",col="#0067E6", lwd=2.5)
lines(d_mean_mutM,main="Density of M Values",col="#E50068", lwd=2.5)
legend('topright', legend=c("WT", "MUT"), 
       fill = c("#0067E6","#E50068"), cex=1
)
```

Overall, by looking at the two graphs seems that WT and MUT have a similar distribution for both the beta value and the M value. However, there are some differences: - *central values*: WT are slightly lower - *peaks values*: WT are slightly higher

## 7. Functional normalization

Normalizion of the data using the **Functional Normalization** and comparison between raw data and normalized data.

Functional normalization extends the idea of quantile normalization by **adjusting for known covariates measuring unwanted variation**. For the 450k array, the first k principal components of the internal control probes matrix play the role of the covariates adjusting for technical variation. By choosing covariates that only measure unwanted technical variation, functional normalization will only remove the variation explained by these technical measurements and will **leave biological variation intact**. Functional normalization allows a sensible trade-off between not removing any technical variation at all (no normalization) and removing too much variation, including global biological variation, as can occur in quantile normalization.

The output is a plot with 6 panels in which, for both raw and normalized data, there are: density plots of beta mean values according to the chemistry of the probes; density plot of beta standard deviation values according to the chemistry of the probes; boxplot of beta values.

```{r message=FALSE,warning=FALSE}
#creation of two dataframes, containing only type I (dfI) or type II (dfII) probes
dfI <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type=="I",]
dfI <- droplevels(dfI)
dfII <- Illumina450Manifest_clean[Illumina450Manifest_clean$Infinium_Design_Type=="II",]
dfII <- droplevels(dfII)

#subsetting of the beta matrix according to the chemistry of the probes
beta <- getBeta(MSet.raw)
beta_I <- beta[rownames(beta) %in% dfI$IlmnID,]
beta_II <- beta[rownames(beta) %in% dfII$IlmnID,]

#calculation of the mean, sd and their corresponding density for the raw data
mean_of_beta_I <- apply(beta_I,1,mean,na.rm=T)
mean_of_beta_II <- apply(beta_II,1,mean,na.rm=T)
d_mean_of_beta_I <- density(mean_of_beta_I,na.rm=T)
d_mean_of_beta_II <- density(mean_of_beta_II,na.rm=T)

sd_of_beta_I <- apply(beta_I,1,sd,na.rm=T)
sd_of_beta_II <- apply(beta_II,1,sd,na.rm=T)
d_sd_of_beta_I <- density(sd_of_beta_I,na.rm=T)
d_sd_of_beta_II <- density(sd_of_beta_II,na.rm=T)

#application of the Functional Normalization
preprocessFunnorm_results <- preprocessFunnorm(RGset)
beta_Funnorm <- getBeta(preprocessFunnorm_results)
beta_Funnorm_I <- beta_Funnorm[rownames(beta_Funnorm) %in% dfI$IlmnID,]
beta_Funnorm_II <- beta_Funnorm[rownames(beta_Funnorm) %in% dfII$IlmnID,]


#calculation of the mean, sd and their corresponding density for the normalized data
mean_of_beta_Funnorm_I <- apply(beta_Funnorm_I,1,mean,na.rm=T)
mean_of_beta_Funnorm_II <- apply(beta_Funnorm_II,1,mean,na.rm=T)
d_mean_of_beta_Funnorm_I <- density(mean_of_beta_Funnorm_I,na.rm=T)
d_mean_of_beta_Funnorm_II <- density(mean_of_beta_Funnorm_II,na.rm=T)

sd_of_beta_Funnorm_I <- apply(beta_Funnorm_I,1,sd,na.rm=T)
sd_of_beta_Funnorm_II <- apply(beta_Funnorm_II,1,sd,na.rm=T)
d_sd_of_beta_Funnorm_I <- density(sd_of_beta_Funnorm_I,na.rm=T)
d_sd_of_beta_Funnorm_II <- density(sd_of_beta_Funnorm_II,na.rm=T)
```

```{r, fig.align = 'center',fig.height=10, fig.width=20, message=FALSE, warning=FALSE}
par(mfrow=c(2,3))
targets$Group <- as.factor(targets$Group)
palette(c("#E50068", "#0067E6"))

#Raw data graphs
plot(d_mean_of_beta_I,col="#77FF00",main="raw Beta density", lwd=2.5)
lines(d_mean_of_beta_II,col="#FF7038",lwd=2.5)
legend('topright', legend=c("Type I","Type II"), fill = c("#77FF00","#FF7038"), cex=1)
plot(d_sd_of_beta_I,col="#77FF00",main="raw sd density", lwd=2.5)
lines(d_sd_of_beta_II,col="#FF7038",lwd=2.5)
legend('topright', legend=c("Type I","Type II"), fill = c("#77FF00","#FF7038"), cex=1)
boxplot(beta, col = targets$Group)
title('Boxplot of raw Beta values')


#Normalized data graphs
plot(d_mean_of_beta_Funnorm_I,col="#77FF00",main="preprocessFunnorm Beta density",lwd=2.5)
lines(d_mean_of_beta_Funnorm_II,col="#FF7038",lwd=2.5)
legend('topright', legend=c("Type I","Type II"), fill = c("#77FF00","#FF7038"), cex=1)
plot(d_sd_of_beta_Funnorm_I,col="#77FF00",main="preprocessFunnorm sd density",lwd=2.5)
lines(d_sd_of_beta_Funnorm_II,col="#FF7038",lwd=2.5)
legend('topright', legend=c("Type I","Type II"), fill = c("#77FF00","#FF7038"), cex=1)
boxplot(beta_Funnorm, col = targets$Group)
title('Boxplot of normalized Beta values')


```

The Functional Normalization produced some differences: the mean density of the raw data showed more difference between the two types of probes, while after the normalization they appear more overlapping. Looking at the boxplot, for the raw data the variation is more visible than for the normalized one. Also, the median and Q1 have been shifted to lower values.

## 8. PCA

Performing PCA on the matrix of normalized beta values generated in step 7, after normalization Checking how samples cluster by group, sex or batch.

```{r, fig.align = 'center',fig.height=7, fig.width=12, message=FALSE, warning=FALSE}
pca_results <- prcomp(t(beta_Funnorm),scale=T)
par(mfrow=c(1,1))


library(factoextra)
fviz_eig(pca_results, addlabels = T,xlab='PC number',ylab='% of variance', barfill = "#0063A6", barcolor = "black")
```

As we can see from the Scree Plot, the **first two PCs** explain the **55.5%** of the variance.

```{r, fig.align = 'center',fig.height=7, fig.width=12, message=FALSE, warning=FALSE}
targets$Group <- as.factor(targets$Group)
palette(c("#E50068", "#0067E6"))
plot(pca_results$x[,1], pca_results$x[,2],cex=1,pch=19,col=targets$Group,xlab="PC1",ylab="PC2",xlim=c(-700,700),ylim=c(-700,700), main=' PCA (Groups)')
text(pca_results$x[,1], pca_results$x[,2],labels=rownames(pca_results$x),cex=0.4,pos=3)
legend("bottomright",legend=levels(targets$Group),col=c(1,2),pch=19, cex=1.0)

```

The groups are divided mainly by the PC1, indeed all the WTs are at lower values of PC1 with respect to the MUTs which are mainly located in the right part of the graph, where the PC1 values are higher. However, while the WTs are well clustered according also to PC2, the same is not for the MUTs which are located at different values of PC2, except for R02C02 and R03C02 which cluster very well. Interesting to notice that R02C01, although being a MUT, clusters with the WTs.

```{r, fig.align = 'center',fig.height=7, fig.width=12, message=FALSE, warning=FALSE}
targets$Sex <- as.factor(targets$Sex)
palette(c("#C364CA", "#8BC4F9"))
plot(pca_results$x[,1], pca_results$x[,2],cex=1,pch=19,col=targets$Sex,xlab="PC1",ylab="PC2",xlim=c(-700,700),ylim=c(-700,700), main='PCA (Sex)')
text(pca_results$x[,1], pca_results$x[,2],labels=rownames(pca_results$x),cex=0.4,pos=3)
legend("bottomright",legend=levels(targets$Sex),col=c(1:nlevels(targets$Group)),pch=19,cex=1.0)


```

By looking at the graph, we can clearly see a division with respect to the sex according to PC1. Moreover, males cluster well together also according to PC2, the same is not for the females. Females are divided into three clusters, which can be found at different levels of PC2. One of the females cluster is really close to the males cluster, especially according to PC2.

```{r, fig.align = 'center',fig.height=7, fig.width=12, message=FALSE, warning=FALSE}
targets$Slide <- as.factor(targets$Slide)
palette(c("#9e0059", "green"))
plot(pca_results$x[,1], pca_results$x[,2],cex=1,pch=19,col=targets$Slide,xlab="PC1",ylab="PC2",xlim=c(-700,700),ylim=c(-700,700), main=' PCA (Batch)')
text(pca_results$x[,1], pca_results$x[,2],labels=rownames(pca_results$x),cex=0.4,pos=3)
legend("bottomright",legend=levels(targets$Slide),col=c(1,2),pch=19, cex=1.0)
```

Since all the samples belong to the same batch (**200400320115**) we cannot see a difference in terms of clustering.

## 9. Differential Methylation Analysis

Identification of differentially methylated probes between group WT and group MUT through the application of the Mann Whitney test.

```{r, message=FALSE, warning=FALSE}
library(future.apply) #for parallelization
plan(multisession)
My_MannWhitney_function <- function(x) {
  wilcox <- wilcox.test(x~ targets$Group)
  return(wilcox$p.value)
} 
pValues_wilcox <- future_apply(beta_Funnorm,1, My_MannWhitney_function)

final_wilcox<- data.frame(beta_Funnorm, pValues_wilcox)

final_wilcox <- final_wilcox[order(final_wilcox$pValues_wilcox),]

table(final_wilcox$pValues_wilcox)


```

As we can see from the table, the **distribution** of the nominal P-values is **discrete** since we applied a non-parametric test.

## 10. Multiple test correction

Application of multiple test correction, setting a significant threshold of **0.05**.

```{r, message=FALSE, warning=FALSE}
#multiple testing correction
corrected_pValues_BH <- p.adjust(final_wilcox$pValues_wilcox,"BH")
corrected_pValues_Bonf <- p.adjust(final_wilcox$pValues_wilcox,"bonferroni")

#creation of a dataframe that contains the nominal p-values and the p-values after both corrections.
final_wilcox_corrected <- data.frame(final_wilcox, corrected_pValues_BH, corrected_pValues_Bonf)

#differentially methylated probes according to the threshold (both before and after the corrections)
before_correction <- nrow(final_wilcox_corrected[final_wilcox_corrected$pValues_wilcox<= 0.05,])
after_Bonferroni <- nrow(final_wilcox_corrected[final_wilcox_corrected$corrected_pValues_Bonf <= 0.05,])
after_BH <- nrow(final_wilcox_corrected[final_wilcox_corrected$corrected_pValues_BH <= 0.05,])



diff_meth_df <- data.frame(before_correction, after_Bonferroni, after_BH)
rownames(diff_meth_df) <- c("# differentially methylated probes")
diff_meth_df
```

As we can see, both multiple test corrections led to **non significant results**.

```{r, fig.align='center', fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
par(mfrow=c(1,1))
boxplot(final_wilcox_corrected [,9:11], col = c("#7038FF", "#C7FF38", "seashell2"), names=NA)
legend("bottomright", legend=c("raw", "BH", "Bonferroni"),col=c("#7038FF", "#C7FF38", "seashell2"), lty=1:1, cex=0.8, xpd=TRUE, pch=15)
```

## 11. Volcano and Manhattan plots

Production of a Volcano plot and a Manhattan plot of the results of differential methylation analysis

```{r, message=FALSE, warning=FALSE}
# WT group mean
WT_group <- final_wilcox_corrected[,targets$Group=="WT"]
WT_group_mean <- apply(WT_group, 1, mean)

# MUT group mean
MUT_group <- final_wilcox_corrected[,targets$Group=="MUT"]
MUT_group_mean <- apply(MUT_group, 1, mean)

# Compute delta
delta <- WT_group_mean - MUT_group_mean

#create a dataframe
toVolcPlot <- data.frame(delta, -log10(final_wilcox_corrected$pValues_wilcox))
head(toVolcPlot)
```

```{r, fig.align = 'center', message=FALSE, warning=FALSE}
par(mfrow=c(1,1))
HighLight <- toVolcPlot[abs(toVolcPlot[,1])>0.1 & toVolcPlot[,2]>(-log10(0.05)),]
plot(toVolcPlot[,1],toVolcPlot[,2], pch=19, cex=0.6 ,xlab="delta",ylab="-log10(Pvalue)", col="black")
abline(a=-log10(0.05),b=0, col="#fb8b24")
points(HighLight[,1], HighLight[,2], pch=17, cex=0.6, col="#00f5d4")
```

```{r, message=FALSE, warning=FALSE}
library(qqman)
final_wilcox_corrected_df <- data.frame(rownames(final_wilcox_corrected),final_wilcox_corrected)
colnames(final_wilcox_corrected_df)[1] <- "IlmnID"

final_wilcox_corrected_annotated <- merge(final_wilcox_corrected_df, Illumina450Manifest_clean,by="IlmnID")

input_Manhattan <- data.frame(ID=final_wilcox_corrected_annotated$IlmnID,CHR=final_wilcox_corrected_annotated$CHR, MAPINFO=final_wilcox_corrected_annotated$MAPINFO, PVAL=final_wilcox_corrected_annotated$pValues_wilcox)
head(input_Manhattan)

levels(input_Manhattan$CHR)[levels(input_Manhattan$CHR) == "X"] <- "23"               
levels(input_Manhattan$CHR)[levels(input_Manhattan$CHR) == "Y"] <- "24"              
input_Manhattan$CHR <- as.numeric(as.character(input_Manhattan$CHR))

```

```{r, fig.align='center', fig.height=8, fig.width=16, message=FALSE, warning=FALSE}
manhattan(input_Manhattan, snp="ID",chr="CHR", bp="MAPINFO", p="PVAL", suggestiveline = -log10(0.05),col=rainbow(24), annotateTop = F)
?manhattan

```

*In both the Volcano and the Manhattan plot the data are "stratified", because of the application of a non-parametric test.*

## 12. Heatmap

Production of an heatmap of the top 100 differentially methylated probes

```{r, fig.align = 'center',fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
library(gplots)
library(viridis)
input_heatmap=as.matrix(final_wilcox[1:100,1:8])

```

#### Linkage method: complete

```{r, fig.align = 'center',fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
group_color = c()
i = 1
for (name in colnames(beta)){
  if (name %in% wt){group_color[i]="#0067E6"}
  else{group_color[i]="#E50068"}
  i = i+1
}

heatmap.2(input_heatmap,col=viridis(100),Rowv=T,Colv=T,dendrogram="both",key=T,ColSideColors=group_color,density.info="none",trace="none",scale="none",symm=F,main="Complete linkage",key.xlab='beta-val',key.title=NA,keysize=1,labRow=NA)
legend("topright", legend=levels(targets$Group),col=c('#E50068','#0067E6'),pch = 19,cex=0.7)
```

#### Linkage method: single

```{r, fig.align = 'center',fig.height=7, fig.width=7, message=FALSE, warning=FALSE}

heatmap.2(input_heatmap,col=viridis(100),Rowv=T,Colv=T,hclustfun = function(x) hclust(x,method = 'single'),dendrogram="both",key=T,ColSideColors=group_color,density.info="none",trace="none",scale="none",symm=F,main="Single linkage",key.xlab='beta-val',key.title=NA,keysize=1,labRow=NA)
legend("topright", legend=levels(targets$Group),col=c('#E50068','#0067E6'),pch = 19,cex=0.7)
```

#### Linkage method: average

```{r, fig.align = 'center',fig.height=7, fig.width=7}

heatmap.2(input_heatmap,col=viridis,Rowv=T,Colv=T,hclustfun = function(x) hclust(x,method = 'average'),dendrogram="both",key=T,ColSideColors=group_color
          ,density.info="none",trace="none",scale="none",symm=F,main="Average linkage",key.xlab='beta-val',key.title=NA,keysize=1,labRow=NA)
legend("topright", legend=levels(targets$Group),col=c('#E50068','#0067E6'),pch = 19,cex=0.7)
```
